Sequential rank agreement methods for comparison of lists

### 2015-08-06 latex/paper.Rnw
Minor tweaks

The MALOVA study is a multidisciplinary Danish study on ovarian cancer \citep{Hogdall:2004:Cancer:15160342} where all Danish women diagnosed with an ovarian tumor and referred for surgery from the participating departments of gynecology were enrolled continuously from December 1994 to May 1999. For the purpose of illustration we use a random subsample of $119$ patients with a total of $58$ patients with malignant ovarian cancers as cases and $61$ patients with benign
ovarian tumors as controls. The DACOVA study is another multidisciplinary Danish study on ovarian cancer which included about $66\%$ of the female population of Denmark \citep{bertelsen1991protocol}. It was aimed to continuously enroll all patients that were referred to surgery of an ovarian tumor clinically suspected to be cancer during the period 1984 to 1990. Similarly, we use a random subsample of $113$ patients with a total of $54$ malignant ovarian cancers and $59$ benign ovarian tumors/gynecologic disorders.

@@

We use the following four classification methods: Random Forest \citep{breiman2001random} implemented in the R package \texttt{randomForest} \citep{liaw2002classification}, logistic Lasso \citep{tibshirani1996regression} and Ridge regression \citep{segerstedt1992ordinary} both implemented in the R package \texttt{glmnet} \citep{friedman2010regularization}, and Partial Least Squares Discriminant Analysis (PLS-DA) \citep{boulesteix2004pls} implemented in the R package \texttt{caret} \citep{Jed-Wing:2014aa}.

In both scenarios we use the MALOVA data to train the statistical models, and the two types of agreements are assessed with respect to perturbations of the training data in the following manner. We repeatedly draw a sub-sample (without replication) consisting of 90\% of the MALOVA observations and train the four models on the sub-sample. Each of the models produces a ranking of the 5000 predictors according to their importance with respect to discriminating between  the tumor types, and the agreement of these rankings across the different sub-samples are used to assess the stability of ranking the importance of the predictors. For the Random Forest classifier the predictors are ranked according to the Gini index, while for the logistic Lasso and Ridge regression models we order the predictors by absolute magnitude of the estimated regression coefficients. For the PLS-DA model the importance of the predictors is based on a weighted sum of the absolute coefficients where the weights are proportional to the reduction in the sums of squares across the components. To assess the stability of the individual risk predictions we apply the predictors from the DACOVA data set to each of the models. These predicted probabilities are then ranked in decreasing order, and the agreement of the risk predictions are similarly obtained by the sequential rank agreement applied across the different sub-samples of the training data. We use 1000 iterations for the sub-sampling procedure.

All four models depend on tuning parameters. We estimate these by a 20 times repeated 5-fold cross-validation on the training data in each iteration. The implementation of Lasso and Ridge regression in the \texttt{glmnet} package offers three different cross-validated optimization criteria for the penalty parameter: total deviance, classification accuracy and Area Under ROC. We apply all three criteria to our data to investigate their effect on the agreement.

@@

Figure \ref{fig:app1} (left) shows the sequential rank agreement applied to the estimated risk probabilities of the $113$ patients in the DACOVA data set obtained from the $1000$ sub-samples of the MALOVA data, and Figure \ref{fig:app1} (right) shows the sequential rank agreement of the estimated importance of the 5000 predictors. For clarity we only show the agreement for the latter up to list depth 600 where all the curves have approximately become stable. We note that the predictor importances estimated with the Lasso model are capable of producing censored lists depending on the value of the penalty parameter. Under the deviance optimization criterion the median number of non-zero coefficients was 33 (range 18 to 52). In order to calculate the sequential rank agreement we therefore applied Algorithm \ref{sra-algorithm} with $B = 1000$ and $P=5000$. The effect of the censoring caused by the sparsity of the Lasso is clearly visible in the agreement curves.

>>>>

The MALOVA study is a multidisciplinary Danish study on ovarian cancer
\citep{Hogdall:2004:Cancer:15160342} where all Danish women diagnosed
with an ovarian tumor and referred for surgery from the participating
departments of gynecology were enrolled continuously from December
1994 to May 1999. For the purpose of illustration we use a random
subsample of $119$ patients with a total of $58$ patients with
malignant ovarian cancers as cases and $61$ patients with benign
ovarian tumors as controls. The DACOVA study is another
multidisciplinary Danish study on ovarian cancer which included about
$66\%$ of the female population of Denmark
\citep{bertelsen1991protocol}. The study aimed to continuously enroll
all patients that were referred to surgery of an ovarian tumor
clinically suspected to be cancer during the period from 1984 to
1990. Similarly, we use a random subsample from the DACOVA study of
$113$ patients with a total of $54$ malignant ovarian cancers and $59$
benign ovarian tumors/gynecologic disorders.

@@

Four classification methods are considered: Random Forest
\citep{breiman2001random} implemented in the R package
\texttt{randomForest} \citep{liaw2002classification}, logistic Lasso
\citep{tibshirani1996regression} and Ridge regression
\citep{segerstedt1992ordinary} both implemented in the R package
\texttt{glmnet} \citep{friedman2010regularization}, and Partial Least
Squares Discriminant Analysis (PLS-DA) \citep{boulesteix2004pls}
implemented in the R package \texttt{caret} \citep{Jed-Wing:2014aa}.

In both scenarios we use the MALOVA data to train the statistical
models, and 
%the two types of 
in both situations the
agreements are assessed with respect to
perturbations of the training data in the following manner. We
repeatedly draw a sub-sample (without replication) consisting of 90\%
of the MALOVA observations and train the four models on the
sub-sample. Each of the models produces a ranking of the 5000
predictors according to their importance with respect to
discriminating between the tumor types, and the agreement of these
rankings across the different sub-samples are used to assess the
stability of ranking the importance of the predictors. For the Random
Forest classifier the predictors are ranked according to the Gini
index, while for the logistic Lasso and Ridge regression models we
order the predictors by absolute magnitude of the estimated regression
coefficients. For the PLS-DA model the importance of the predictors is
based on a weighted sum of the absolute coefficients where the weights
are proportional to the reduction in the sums of squares across the
components. To assess the stability of the individual risk predictions
we apply the predictors from the DACOVA data set to each of the
models. These predicted probabilities are then ranked in decreasing
order, and the agreement of the risk predictions are similarly
obtained by the sequential rank agreement applied across the different
sub-samples of the training data. We use 1000 iterations for the
sub-sampling procedure.

All four models depend on tuning parameters. We estimate these by a 20
times repeated 5-fold cross-validation on the training data in each
iteration. The implementation of Lasso and Ridge regression in the
\texttt{glmnet} package offers three different cross-validated
optimization criteria for the penalty parameter: total deviance,
classification accuracy and area under ROC. We apply all three
criteria to our data to investigate their effect on the agreement.

@@

Figure \ref{fig:app1} (left) shows the sequential rank agreement
applied to the estimated risk probabilities of the $113$ patients in
the DACOVA data set obtained from the $1000$ sub-samples of the MALOVA
data, and Figure \ref{fig:app1} (right) shows the sequential rank
agreement of the estimated importance of the 5000 predictors. For
clarity we only show the agreement for the latter up to list depth 600
where all the curves have approximately become stable.  The Lasso
models produce censored lists depending on the value of the penalty
parameter. Under the deviance optimization criterion the median number
of non-zero coefficients was 33 (range 18 to 52). In order to
calculate the sequential rank agreement we therefore applied Algorithm
\ref{sra-algorithm} with $B = 1000$ and $P=5000$. The effect of the
censoring caused by the sparsity of the Lasso is clearly visible in
the agreement curves.
